### 2025-06-20 docs/thesis/main.tex
Add thesis
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\documentclass[noacm,sigplan,review]{acmart}

% \acmSubmissionID{PAPER ID}
\renewcommand\footnotetextcopyrightpermission[1]{}
\settopmatter{printfolios=false,printacmref=false}

\usepackage{setspace}
\usepackage{enumerate}
\usepackage{algorithm2e}
\usepackage{algpseudocode}
\usepackage{graphics}
\usepackage{xparse} 
\usepackage{xspace}
\usepackage{multirow}
\usepackage{csvsimple}
\usepackage{balance}
\usepackage{minted}
\usepackage{xcolor}
\usepackage{nicefrac}
\usepackage{siunitx}
\usepackage{array,framed}
\usepackage{booktabs}
\usepackage{color}
\usepackage{soul}
\usepackage{float}
\usepackage{epsfig}
\usepackage{wrapfig}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{adjustbox}
\usepackage{csquotes}
\usepackage{cleveref}
\usepackage{dirtytalk}
\usepackage{textgreek}
\usepackage{oplotsymbl}

\def\eg{{\em e.g.}, }
\def\ie{{\em i.e.}, }
\def\etc{{\em etc.}\xspace}
\def\vs{{\em vs.}\xspace}

\def\gptmodel{{GPT-4o}\xspace}

\newcommand{\todo}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\sys}{{\scshape Kv{\textalpha}sir}\xspace}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\sx}[1]{(\S\ref{#1})}
\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\se}[1]{\S\ref{#1}}
\newcommand{\fg}[1]{Fig.~\ref{#1}}
\newcommand{\heading}[1]{\vspace{2pt}\noindent\textbf{\emph{#1}}:\enspace}
\newcommand{\ttt}[1]{\texttt{#1}\xspace}
\newcommand{\xxx}{\colorbox{red!30}{xxx}\xspace}
\newcommand{\promptbox}[1]{\hspace*{-0.0135\columnwidth}\fbox{\parbox[t]{0.95\columnwidth}{\tt{#1}}}}

\crefformat{section}{\S#2#1#3}
\crefmultiformat{section}{\S#2#1#3}{--\S#2#1#3}{, \S#2#1#3}{ and \S#2#1#3}

@@

\title{\sys: Controllable Program Regeneration}
\author{Evangelos Lamprou}

% Use cases for something like this include:
% - program repair
% - turning insecure code into secure code by removing side-effects
% - transforming a program in language A to language B
% - turning a program into a more idiomatic version of the same language
% - having a program use a different API
% - have a program be more amendable to parallelization
% - have a program be more amendable to further analysis or transformation

\begin{abstract}
Modern software systems are constantly evolving—requiring developers to
  refactor legacy code, adapt to changing APIs, translate across languages, and
  patch security vulnerabilities. These transformations are often brittle,
  manual, and error-prone. Large language models (LLMs) offer new opportunities
  for automating such tasks, but their outputs are shaped by surface cues, lack
  semantic guarantees, and require brittle prompting strategies. This paper
  presents \sys, a system for controllable program regeneration that combines
  logic-based planning with LLM-guided synthesis.
  By expressing transformation
  goals as declarative constraints over program properties---such as input/output
  behavior, side-effect freedom, language, or modular structure, \sys
  regenerates code that satisfies explicit semantic requirements.
  We demonstrate \sys across a wide range of tasks, including security
  hardening, language translation, and idiomatic rewriting, and highlight its
  ability to deliver interpretable, high-quality transformations with minimal
  developer guidance.
\end{abstract}
\maketitle

\section{Introduction}
\begin{figure}[t]
  % https://docs.google.com/drawings/d/1LAmdVYjfAID24eSM-hl5PRhVg9fRMJMWYETsvvcyAuU/edit?usp=sharing
  \includegraphics[width=\columnwidth]{figs/kvasir_overview.pdf}
  \caption{\textbf{\sys overview}
Given a transformation query, and a source program, \sys extracts a minimal set
  of properties,
  guided by a logic engine and knowledge base.
  It then synthesizes a new program $P'$ that satisfies the query, verifying
  that wanted properties are present and unwanted properties are absent in $P'$.
}
  \label{fig:overview}
\end{figure}

@@

% TODO: Add section references
\heading{Contributions}
The paper's contributions are:
\begin{itemize}
  \item A property-aware transformation framework, combining logic-based reasoning, and LLM-assisted program synthesis~(\cref{sec:design});
 \item A declarative interface and accompanying DSL for program transformation, allowing users to express semantic goals as logical queries over properties of the regenerated program~(\cref{sec:dsl});
 \item A verification architecture that checks synthesized programs against the specified properties, providing feedback for refinement~(\cref{sec:verification});
 \item An empirical evaluation demonstrating \sys's effectiveness and scalability across transformation scenarios~(\cref{sec:evaluation}).
\end{itemize}

@@

\section{Example}
\label{sec:example}
\sys enables users to specify transformation goals declaratively.
Logical constraints over properties of the
regenerated program express these goals. Examples are input-output equivalence, side-effect elimination,
target language, lines of code, and others.

@@

This section shows the application of \sys
to three distinct program transformation tasks, 
targeting security, language translation, compatibility,
idiomaticity, and modularity.

@@

\section{\sys Design}
\label{sec:design}

@@

\section{Property Language and Extraction}
\label{sec:dsl}

@@

\section{Synthesis Plan}
\label{sec:synthesis}

@@

\section{Property Verification and Feedback Loop}
\label{sec:verification}

@@

\section{Evaluation}
\label{sec:evaluation}

@@

\subsection{Benchmarks \& Methodology}

@@

\begin{table}[h]
\centering
  \caption{\textbf{Benchmark summary}. 
  Benchmark programs come from online repositories, public archives, security research datasets, and often referenced examples in programming culture.
  The $N$ column shows the number of programs/functions from each benchmar.
  }
\begin{tabular}{llrl}
\toprule
Benchmark                          & Description         & $N$ & Source \\
\midrule
Rosetta Code                       & Programming tasks   & 16   & \cite{rosettacode} \\
npm utilities                      & Utilities from npm  & 10  & \cite{regbench2025} \\
PyPi utilities                   & Utilities from PyPi & 3   & \cite{regbench2025} \\
SSCA                               & Sneaky SSAs         & 3   & \cite{ev:eurosec:2022, ohm2020backstabber,copeland2019frightening} \\
IOCCC                              & Obfuscated C code   & 2   & \cite{ioccc} \\
Microbenchmarks                    &                     & 3   & \\
\hspace{.5em} \ttt{flatmap-stream} & Secure regeneration &     & \cite{es1}  \\
\hspace{.5em} \ttt{Q\_rsqrt}       & C refactoring       &     & \cite{fast_inv_sqrt}  \\
\hspace{.5em} \textsf{rigidDB}     & Modularization      &     & \cite{codewithsadeemusicplayer} \\
\bottomrule
\end{tabular}
\label{tab:benchmarks}
\end{table}

\subsection{Q1: Correctness and Regeneration Quality}

@@

\subsection{Q2: Naive LLM vs. \sys}

@@

\subsection{Q3: Planning Efficiency \& Precision}

@@

\section{Related Work}

@@

\section{Discussion and Limitations}
\label{sec:discussion}

@@

\section{Conclusion}
This paper introduced \sys, a system for property-guided program regeneration
that bridges the flexibility of large language models with the precision of
logic-driven planning.
By allowing developers to declaratively specify
transformation goals as logical constraints, \sys elevates LLM-assisted program
trasnformations to a controllable, interpretable, and iterative synthesis-verification
process.
Evaluation demonstrates that \sys produces high-quality regenerations
across diverse tasks, outperforming naive prompting baselines in correctness,
controllability, and resilience to adversarial or obfuscated inputs.

\sys opens the door to a new class of synthesis tools that combine symbolic
reasoning and neural generation under a unified, extensible framework.
As program synthesis systems grow increasingly reliant on LLMs, we believe the
future lies in harnessing their generative power within a framework that retains
correctness, explainability, and developer control as first-class citizens.

\bibliographystyle{ACM-Reference-Format}
\bibliography{bib.bib}

\appendix

\section{A Primer on Answer Set Programming (ASP)}
\label{sec:asp}

\emph{Answer Set Programming (ASP)} is a declarative logic programming paradigm
that generalizes Datalog by supporting nonmonotonic reasoning,
negation-as-failure, and optimization. \sys uses ASP to represent and solve
regeneration planning tasks by encoding properties, constraints, and goals as
logic rules.

\heading{From Datalog to ASP}
Like Datalog, ASP programs consist of rules of the form:
\begin{verbatim}
head :- body.
\end{verbatim}

But unlike Datalog
ASP supports \emph{negation-as-failure} using \texttt{not}
ASP allows \emph{disjunctions} and \emph{constraints}
ASP evaluates programs under the \emph{stable model semantics}, which may yield
multiple answer-sets (models), each representing a consistent solution

\heading{Example: Feature Selection}
\begin{verbatim}
{ select(a); select(b); select(c) }.

:- select(a), select(b).   % a and b are incompatible

#maximize { X : select(X) }.   % Prefer selecting more features
\end{verbatim}

This program searches over subsets of \texttt{\{a,b,c\}}, excludes those
violating constraints, and returns models with maximal selections.

>>>>

\documentclass[a4paper,twoside,12pt]{report} %openright
\input{setup/statics.tex}
\input{setup/preamble.tex}
\input{setup/settings.tex}

@@

\pagenumbering{roman}
\input{frontmatter/frontpage.tex}
\pagecolor{white}
\newgeometry{top=2.81cm, bottom=2.75cm, outer=2.5cm, inner=3.5cm}
\pagestyle{empty}
\cleardoublepage 
\input{frontmatter/copyright.tex}
\clearpage 
\pagestyle{main}
\input{frontmatter/approval.tex}
\clearpage 
\input{frontmatter/abstract.tex}
\clearpage 
\input{frontmatter/acknowledgements.tex}
\cleardoublepage 
\tableofcontents
\cleardoublepage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pagenumbering{arabic}

\chapter{Introduction}
\begin{figure}[t]
  % https://docs.google.com/drawings/d/1LAmdVYjfAID24eSM-hl5PRhVg9fRMJMWYETsvvcyAuU/edit?usp=sharing
  \includegraphics[width=\columnwidth]{figs/kvasir_overview.pdf}
  \caption{\textbf{\sys overview}
Given a transformation query, and a source program, \sys extracts a minimal set
  of properties,
  guided by a logic engine and knowledge base.
  It then synthesizes a new program $P'$ that satisfies the query, verifying
  that wanted properties are present and unwanted properties are absent in $P'$.
}
  \label{fig:overview}
\end{figure}

@@

% TODO: Add chapter references
\heading{Contributions}
The paper's contributions are:
\begin{itemize}
  \item A property-aware transformation framework, combining logic-based reasoning, and LLM-assisted program synthesis~(\cref{sec:design});
 \item A declarative interface and accompanying DSL for program transformation, allowing users to express semantic goals as logical queries over properties of the regenerated program~(\cref{sec:dsl});
 \item A verification architecture that checks synthesized programs against the specified properties, providing feedback for refinement~(\cref{sec:verification});
 \item An empirical evaluation demonstrating \sys's effectiveness and scalability across transformation scenarios~(\cref{sec:evaluation}).
\end{itemize}

@@

\chapter{Example}
\label{sec:example}
\sys enables users to specify transformation goals declaratively.
Logical constraints over properties of the
regenerated program express these goals. Examples are input-output equivalence, side-effect elimination,
target language, lines of code, and others.

@@

This chapter shows the application of \sys
to three distinct program transformation tasks, 
targeting security, language translation, compatibility,
idiomaticity, and modularity.

@@

\chapter{\sys Design}
\label{sec:design}

@@

\chapter{Property Language and Extraction}
\label{sec:dsl}

@@

\chapter{Synthesis Plan}
\label{sec:synthesis}

@@

\chapter{Property Verification and Feedback Loop}
\label{sec:verification}

@@

\chapter{Evaluation}
\label{sec:evaluation}

@@

\section{Benchmarks \& Methodology}

@@

\begin{table}[h]
\centering
  \caption{\textbf{Benchmark summary}. 
  Benchmark programs come from online repositories, public archives, security research datasets, and often referenced examples in programming culture.
  The N column shows the number of programs/functions from each benchmar.
  }
\begin{tabular}{llrl}
\toprule
Benchmark                          & Description         & N & Source \\
\midrule
Rosetta Code                       & Programming tasks   & 16   & \cite{rosettacode} \\
npm utilities                      & Utilities from npm  & 10  & \cite{regbench2025} \\
PyPi utilities                   & Utilities from PyPi & 3   & \cite{regbench2025} \\
SSCA                               & Sneaky SSAs         & 3   & \cite{ev:eurosec:2022, ohm2020backstabber,copeland2019frightening} \\
IOCCC                              & Obfuscated C code   & 2   & \cite{ioccc} \\
Microbenchmarks                    &                     & 3   & \\
\hspace{.5em} \ttt{flatmap-stream} & Secure regeneration &     & \cite{es1}  \\
\hspace{.5em} \ttt{Q\_rsqrt}       & C refactoring       &     & \cite{fast_inv_sqrt}  \\
\hspace{.5em} \textsf{rigidDB}     & Modularization      &     & \cite{codewithsadeemusicplayer} \\
\bottomrule
\end{tabular}
\label{tab:benchmarks}
\end{table}

\section{Q1: Correctness and Regeneration Quality}

@@

\section{Q2: Naive LLM vs. \sys}

@@

\section{Q3: Planning Efficiency \& Precision}

@@

\chapter{Related Work}

@@

\chapter{Discussion and Limitations}
\label{sec:discussion}

@@

\chapter{Conclusion}
This paper introduced \sys, a system for property-guided program regeneration
that bridges the flexibility of large language models with the precision of
logic-driven planning.
By allowing developers to declaratively specify
transformation goals as logical constraints, \sys elevates LLM-assisted program
trasnformations to a controllable, interpretable, and iterative synthesis-verification
process.
Evaluation demonstrates that \sys produces high-quality regenerations
across diverse tasks, outperforming naive prompting baselines in correctness,
controllability, and resilience to adversarial or obfuscated inputs.

\sys opens the door to a new class of synthesis tools that combine symbolic
reasoning and neural generation under a unified, extensible framework.
As program synthesis systems grow increasingly reliant on LLMs, we believe the
future lies in harnessing their generative power within a framework that retains
correctness, explainability, and developer control as first-class citizens.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\printbibliography[heading=bibintoc,title={Bibliography}]
\cleardoublepage 
\appendix
\input{backmatter/appendix.tex}
\cleartoleftpage
\input{backmatter/backpage.tex}
